Studying the Performance of the Jellyfish Search Optimiser for the Application of Projection Pursuit

### 2024-06-14 optim.tex
update the squintability calculation and misc

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{algorithm}
\usepackage{algpseudocode}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Journal of Multivariate Analysis}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num-names}
\usepackage{bookmark}

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, rightrule=.15mm, opacitybacktitle=0.6, colframe=quarto-callout-note-color-frame, opacityback=0, bottomrule=.15mm, titlerule=0mm, breakable, left=2mm, colback=white, toptitle=1mm, bottomtitle=1mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, toprule=.15mm, coltitle=black, colbacktitle=quarto-callout-note-color!10!white, arc=.35mm]

@@

The data structure proposed in \citet{RJ-2021-105} is implemented for JS
optimisers and additional information such as index function used, data
dimension, and jellyfish hyper-parameters can be included as additional
columns. Below prints an example data collected from finding the
sine-wave structure in 6D data (\texttt{d\ =\ 6}) using a spline index
function(\texttt{splines2d}):

@@

\begin{itemize}
\item
  \textbf{success rate}: Success rate is defined as the proportion of
  simulations that achieve a final index value within 0.05 of the best
  index value found among all 50 simulations. In the dataset above, the
  optimal index value is 0.999. Out of the 50 simulations, 44 achieved
  an index value within {[}0.949, 0.999{]}, resulting in a success rate
  of 0.88. Figure~\ref{fig-projection} shows the best projection found
  in each simulation, sorted by the index value. Most of the projections
  show a clear sine-wave structure, while the last six shows only a
  straight line with clusters at the end. When local optima exist in the
  projection pursuit problem, plotting the projections found in each
  simulation or by each jellyfish can help identify these local optima.
\item
  \textbf{Projection distance}: The projection distance between
  projection bases and the theoretical best basis (or the empirical best
  basis if the theoretical best is unknown) is calculated as the
  Frobenius norm, which measures how close a visited basis is to the
  best basis. Plotting this projection distance against the index value
  for each basis visited reveals how the index value changes as the
  basis approaches the optimal basis, helping to identify the
  squintability of the index function.
  Figure~\ref{fig-index-value-proj-dist} shows the index value against
  the projection distance for the given example and its binned version.
  As the projection distance decreases, the index value initially
  increases with an increasing slope and then level off. The index value
  shows greater variation at medium projection distances compared to
  small and large distances.
\end{itemize}

@@

The JS optimiser provides a new swarm-based search strategy to optimise
projection pursuit problems. Two examples are provided in this section
to demonstrate its performance. The first example compares its
performance with the search better optimiser and explores how it behaves
with different combinations of hyper-parameter. The second example
studies the effect of optimisation properties defined in
Section~\ref{sec-theory}, along with jellyfish hyper-parameters, on the
outcome of the optimisation.

\subsection{Performance of JS optimiser in pipe-finding
problems}\label{sec-app-1}

The performance of JS optimiser is investigated in two folds: 1) compare
it with a commonly used optimiser: the search better optimiser
\citep{RJ-2021-105, laa_using_2020}, and 2) examine its success rate
under different hyper-parameters (number of jellyfishes and the maximum
number of tries). The projection pursuit problem used is finding the
pipe shape using the holes index, investigated by
\citet{laa_using_2020}.

@@

Fifty (50) simulations are conducted with both JS and the search better
optimiser, in four dimensions (\(d = 6, 8, 10, 12\)). The JS optimiser
uses 100 jellyfishes and 100 maximum number of tries, while the search
better optimiser allows a maximum of 1000 samples at each iteration
before the algorithm terminates. Figure~\ref{fig-proj} presents the
final projections found by the two optimisers, broken down by 10th
quantile, faceted by the data dimension. In the 6-dimensional data
scenario, the JS optimiser consistently identifies a clear pipe shape.
The search better optimiser, although also find the pipe shape but with
a wide rim, suggesting a further polish search may be required. With
increasing dimensions, the JS optimiser may not always identify the pipe
shape due to random sampling, it still presents a pipe shape in over
50\% of cases. When compared to the search better optimiser, the JS
optimiser reaches higher index values and clearer pipe shapes across all
quantiles.

In the second experiment, fifty (50) simulations are conducted at each
hyper-parameter combination to analyse its effects on the success rate
of the JS optimiser. The hyper-parameters tested include: 1) 20, 50, and
100 jellyfishes, and 2) 50 and 100 maximum number of tries. In each
simulation, success rate is calculated as defined in
Section~\ref{sec-vis}. Figure~\ref{fig-proportion} presents the
proportion of success for each hyper-parameter combinations. As the
number of jellyfish and maximum tries increase, the success rate also
increases. For simpler problems (6 dimensions), small parameter values
(20 jellyfishes and a maximum number of tries of 50) can already result
in high success rate, while for higher-dimensional problems (8, 10, and
12 dimensions), a combination of 100 jellyfishes and 100 maximum number
of tries is necessary for at least half of the jellyfishes to find the
optimal projection.

\subsection{Factors affecting JS optimiser success rate: optimisation
properties and jellyfish hyper-parameters}\label{sec-app-2}

The optimisation properties, including smoothness, squintability, and
speed, offer numerical metrics to characterise the complexity of a
projection pursuit optimisation problem. This example investigates how
these metrics, alongside jellyfish hyper-parameters (the number of
jellyfish and the maximum number of tries), affect the success of the JS
optimiser. Simulations are conducted to obtain the success rate across
various projection pursuit problems, characterised by shape-to-find,
data dimension, and index function used, as well as different
hyper-parameter combinations. Smoothness and squintability are
calculated for each projection pursuit problem as outlined in
Section~\ref{sec-theory}. A generalised linear model is used to
construct the relationship between the success rate, jellyfish
hyper-parameters and optimisation properties.

In addition to the pipe-finding problem, a new shape, sine wave, is
investigated in 6D and 8D spaces with six indices considered:
\texttt{dcor2d\_2}, \texttt{loess2d}, \texttt{MIC}, \texttt{TIC},
\texttt{spline}, and \texttt{stringy}. Combining with two jellyfish
hyper-parameters, a total of 52 cases is produced, comprising of 24
pipe-finding cases and 28 sine-wave finding cases. For each case, the JS
optimiser is run fifty (50) times and the success rate is calculated as
in Section~\ref{sec-app-1}.

\emph{Note: should we print out the 52 cases investigated somewhere, say
appendix, since it is not full grid simulation (otherwise for the
sine-wave problem there will be 6 indexes x 6 configurations = 36 cases
for 6-D data and 2 indexes {[}MIC \& TIC{]} x 6 configurations = 12
cases for 8-D data)? We only do a selection of 28 of them to save time/
computational resources.}

Smoothness and squintability are computed for each case, following the
procedures outlined in Section~\ref{sec-smoothness} and
Section~\ref{sec-squintability}. To calculate smoothness, three hundred
random bases are simulated. Index values and projection distance (to the
optimal basis) are calculated for each random basis before fitting them
into a Gaussian process model. For squintability, fifty random bases are
sampled and interpolated to the optimal basis with a step size of 0.005.
For these interpolated bases, index values and projection distances to
the optimal basis are calculated and averaged with a bin width of 0.005.
The squintability measure is then calculated from fitting a 4-parameter
scaled logistic function of index value against projection distance,
estimated by non-linear least square.

\emph{Note: Shall I repeat the scaled logistic function with our
parameterisation here? It can be handy to interpret theta1 - theta4 but
they should've already been explained in the squintability section.}

Table~\ref{tbl-smoothness-squintability} presents the parameters
estimated from the Gaussian process (variance, range, smooth, and
nugget) and the scaled logistic function (\(\theta_1\) to \(\theta_4\))
for each case. The column ``smooth'' is used as the smoothness measure
and the column ``squint'' is calculated as
\(\frac{\theta_1 \theta_2 \theta_3}{4}\) as squintability. The low
squint value of MIC and TIC index is due to its convex shape of the
index value against the projection distance, as shown in
Figure~\ref{fig-idx-proj-dist}. Comparing to other concave shapes, the
maximum first derivative happens at a smaller projection distance
(\(\theta_2\)), requiring the optimiser to get closer to the optimal to
see a significant change in the index value, hence a small squintability
measure.

\emph{Note: the stringy squintability doesn't make sense}

Table~\ref{tbl-mod-data} combines all the variables calculated above
(the JS optimiser success rate, jellyfish hyper-parameters, and
optimisation properties) into one table. Three pre-processing steps are
applied to the data: 1) scaling the jellyfish hyper-parameters by a
factor of 10 for interpretation, 2) creating a new binary variable
\texttt{long\_time} to indicate cases with an average run time over 30
seconds, and 3) encoding the success rate for the stringy case as 0,
since none of the 50 simulations identified the sine-wave shape. A
generalised linear model with a binomial family and a logit link
function is used to fit the data and Table~\ref{tbl-mod-output} presents
the model outputs. The model suggests that the success rate of the JS
optimiser is positively associated with the two jellyfish
hyper-parameters, as well as with smoothness and squintability. However,
being flagged with long runtime and an increase of data dimension will
reduce the success rate. The variable \texttt{squintability} and
\texttt{dimension} are significant, suggesting their importance relative
to jellyfish hyper-parameters in the optimisation success.

\begin{longtable}[t]{|>{}ll>{}l|rrr>{}r|rrrr>{}r|>{}r|}

\caption{\label{tbl-smoothness-squintability}Parameters estimated from
the Gaussian process (including variance, range, smoothness, and nugget)
and scaled logistic function (\(\theta_1\) to \(\theta_4\)) for the
pipe-finding and sine-wave finding problems. The ``smooth'' column and
``squint'' column, calculated as
\(\frac{\theta_1 \theta_2 \theta_3}{4}\), represent the smoothness and
squintability measures.}

\toprule
 & shape & index & d & variance & range & smooth & nugget & theta1 & theta2 & theta3 & theta4 & squint\\
\midrule
1 & pipe & holes & 6 & 0.002 & 0.408 & 2.364 & 0.212 & 0.991 & 0.860 & 3.368 & 0.172 & 0.718\\
2 & pipe & holes & 8 & 0.000 & 0.259 & 2.373 & 0.613 & 1.011 & 0.869 & 3.264 & 0.085 & 0.716\\
3 & pipe & holes & 10 & 0.000 & 0.144 & 2.317 & 1.831 & 1.011 & 0.885 & 3.151 & 0.153 & 0.705\\
4 & pipe & holes & 12 & 0.000 & 0.254 & 2.173 & 0.879 & 0.999 & 0.878 & 3.345 & 0.196 & 0.733\\
5 & sine & MIC & 6 & 0.016 & 0.100 & 2.394 & 0.087 & 0.894 & 0.571 & 1.623 & -0.024 & 0.207\\
\addlinespace
6 & sine & MIC & 8 & 0.016 & 0.100 & 2.394 & 0.087 & 0.932 & 0.328 & 1.314 & -0.030 & 0.100\\
7 & sine & TIC & 6 & 0.124 & 0.104 & 2.471 & 0.086 & 0.951 & 0.536 & 1.719 & -0.025 & 0.219\\
8 & sine & TIC & 8 & 0.124 & 0.104 & 2.471 & 0.086 & 0.945 & 0.564 & 1.723 & -0.027 & 0.230\\
9 & sine & dcor2d\_2 & 6 & 0.034 & 0.167 & 2.663 & 0.114 & 0.954 & 1.039 & 2.742 & -0.019 & 0.679\\
10 & sine & loess2d & 6 & 0.083 & 0.307 & 2.194 & 0.292 & 1.016 & 1.039 & 2.648 & 0.080 & 0.699\\
\addlinespace
11 & sine & splines2d & 6 & 0.040 & 0.189 & 2.606 & 0.104 & 1.014 & 1.051 & 2.730 & -0.009 & 0.727\\
12 & sine & stringy & 6 & 0.000 & 1173.035 & 1.031 & 17608.047 & 1.040 & 0.011 & 254.733 & 0.170 & 0.731\\
\bottomrule

@@

\toprule
Index & D & Smoothness & Squintability & Num. of jellyfish & Max. Num. of tries & Prop. of success & Time\\
\midrule
MIC & 6 & 2.394 & 0.207 & 20 & 50 & 0.12 & 2.479 secs\\
MIC & 6 & 2.394 & 0.207 & 20 & 100 & 0.24 & 8.950 secs\\
MIC & 6 & 2.394 & 0.207 & 50 & 50 & 0.52 & 5.651 secs\\
MIC & 6 & 2.394 & 0.207 & 50 & 100 & 0.64 & 13.223 secs\\
MIC & 6 & 2.394 & 0.207 & 100 & 50 & 0.76 & 19.453 secs\\
\addlinespace
MIC & 8 & 2.394 & 0.100 & 20 & 50 & 0.08 & 2.566 secs\\
MIC & 8 & 2.394 & 0.100 & 20 & 100 & 0.08 & 4.960 secs\\
\bottomrule

@@

\toprule
term & estimate & std.error & statistic & p.value\\
\midrule
Intercept & -4.849 & 4.496 & -1.078 & 0.281\\
Smoothness & 1.672 & 1.527 & 1.094 & 0.274\\
Squintability & 6.967 & 2.173 & 3.207 & 0.001\\
dimension (d) & -0.588 & 0.257 & -2.292 & 0.022\\
long time & -0.816 & 1.333 & -0.612 & 0.540\\
\addlinespace
number of jellyfish & 0.229 & 0.131 & 1.752 & 0.080\\
maximum number of tries & 0.108 & 0.153 & 0.703 & 0.482\\
\bottomrule

@@

\section{Additional plot:}\label{additional-plot}

\caption{\label{fig-idx-proj-dist}Index values against projection
distance for the 12 pipe/sine-wave finding problem after the binning
procedure during the estimation of the squintability measure. The index
value is averaged at each bin width of 0.005 and the \texttt{TIC} index
is scaled to 0-1 for comparison. When finding the sine-wave structure
using the \texttt{MIC} and \texttt{TIC} index, a convex curved is
observed, as opposed to the pipe-finding problem or the sine-wave
finding problem with the \texttt{dcor2d\_2}, \texttt{loess2d}, and
\texttt{splines2d} index. When finding the sine-wave with the stringy
index, the index shows an instantaneous jump to the optimum when close
to the best basis.}

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{algorithm}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\journal{Journal of Multivariate Analysis}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{elsarticle-num-names}
\usepackage{bookmark}

@@

\begin{tcolorbox}[enhanced jigsaw, opacitybacktitle=0.6, arc=.35mm, colbacktitle=quarto-callout-note-color!10!white, titlerule=0mm, toptitle=1mm, colframe=quarto-callout-note-color-frame, title={Algorithm: Jellyfish Optimizer Pseudo Code}, coltitle=black, colback=white, bottomrule=.15mm, rightrule=.15mm, left=2mm, bottomtitle=1mm, breakable, toprule=.15mm, leftrule=.75mm, opacityback=0]

@@

The data structure proposed in \citet{RJ-2021-105} is implemented for JS
optimisers and additional information such as index function used, data
dimension, and jellyfish hyper-parameters can be included as additional
columns. An example data collected from finding the sine-wave structure
in 6D data (\texttt{d\ =\ 6}) using a spline index
function(\texttt{splines2d}) is printed below:

@@

\begin{itemize}
\item
  \textbf{Success rate}: The success rate is defined as the proportion
  of simulations that achieve a final index value within 0.05 of the
  best index value found among all 50 simulations. In the dataset above,
  the optimal index value is 0.999. Out of the 50 simulations, 44
  achieved an index value within {[}0.949, 0.999{]}, resulting in a
  success rate of 0.88. Figure~\ref{fig-projection} shows the best
  projection found in each simulation, sorted by the index value. Most
  of the projections show a clear sine-wave structure, while the last
  six show only a straight line with clusters at the end. When local
  optima exist in the projection pursuit problem, plotting the
  projections found in each simulation or by each jellyfish can help
  identify these local optima.
\item
  \textbf{Projection distance}: The projection distance between
  projection bases and the theoretical best basis (or the empirical best
  basis if the theoretical best is unknown) is calculated as the
  Frobenius norm, which measures how close a visited basis is to the
  best basis. Plotting this projection distance against the index value
  for each basis visited reveals how the index value changes as the
  basis approaches the optimal basis, helping to identify the
  squintability of the index function.
  Figure~\ref{fig-index-value-proj-dist} shows the index value against
  the projection distance for the given example and its binned version.
  As the projection distance decreases, the index value initially
  increases with an increasing slope and then levels off. The index
  value shows greater variation at medium projection distances compared
  to small and large distances.
\end{itemize}

@@

The JS optimiser provides a new swarm-based search strategy to optimise
projection pursuit problems. Two examples are provided in this section
to demonstrate its performance. The first example compares its
performance with the search-better optimiser and explores how it behaves
with different combinations of hyper-parameters. The second example
studies the effect of optimisation properties defined in
Section~\ref{sec-theory}, along with jellyfish hyper-parameters, on the
outcome of the optimisation.

\subsection{Performance of the JS optimiser in pipe-finding
problems}\label{sec-app-1}

The performance of the JS optimiser is investigated in two folds: 1)
compare it with a commonly used optimiser: the search-better optimiser
\citep{RJ-2021-105, laa_using_2020}, and 2) examine its success rate
under different hyper-parameters (number of jellyfishes and maximum
number of tries). The projection pursuit problem used is finding the
pipe shape using the holes index, investigated by
\citet{laa_using_2020}.

@@

Fifty simulations are conducted with both JS and the search-better
optimiser, in four dimensions (\(d = 6, 8, 10, 12\)). The JS optimiser
uses 100 jellyfishes and 100 maximum number of tries, while the
search-better optimiser allows a maximum of 1000 samples at each
iteration before the algorithm terminates. Figure~\ref{fig-proj}
presents the final projections found by the two optimisers, broken down
by 10th quantile, faceted by the data dimension. In the 6-dimensional
data scenario, the JS optimiser consistently identifies a clear pipe
shape. The search-better optimiser also finds the pipe shape but with a
wide rim, suggesting a further polish search may be required. With
increasing dimensions, the JS optimiser may not always identify the pipe
shape due to random sampling, but it still presents a pipe shape in over
50\% of cases. When compared to the search-better optimiser, the JS
optimiser reaches higher index values and clearer pipe shapes across all
quantiles.

In the second experiment, fifty simulations are conducted at each
hyper-parameter combination to analyse its effects on the success rate
of the JS optimiser. The hyper-parameters tested include: 1) 20, 50, and
100 jellyfishes, and 2) 50 and 100 maximum number of tries. In each
simulation, the success rate is calculated as defined in
Section~\ref{sec-vis}. Figure~\ref{fig-proportion} presents the
proportion of success for each hyper-parameter combinations. As the
number of jellyfishes and maximum tries increase, the success rate also
increases. For simpler problems (6 dimensions), small parameter values
(20 jellyfishes and a maximum number of tries of 50) can already result
in high success rate, while for higher-dimensional problems (8, 10, and
12 dimensions), a combination of 100 jellyfishes and 100 maximum number
of tries is necessary for at least half of the jellyfishes to find the
optimal projection.

\subsection{Factors affecting the JS optimiser success rate:
optimisation properties and jellyfish hyper-parameters}\label{sec-app-2}

The optimisation properties, including smoothness, squintability, and
speed, offer numerical metrics to characterise the complexity of a
projection pursuit optimisation problem. This example investigates how
these metrics, along with the jellyfish search hyper-parameters (the
number of jellyfishes and the maximum number of tries), affect the
success of the JS optimiser. Simulations are conducted to obtain the
success rate across various projection pursuit problems, characterised
by shape-to-find, data dimension, and index function used, as well as
different hyper-parameter combinations. Smoothness and squintability are
calculated for each projection pursuit problem as outlined in
Section~\ref{sec-theory}. A generalised linear model is used to
construct the relationship between the success rate, jellyfish
hyper-parameters and optimisation properties.

In addition to the pipe-finding problem, a new shape, sine wave, is
investigated in 6D and 8D spaces with six indices considered:
\texttt{dcor2d\_2}, \texttt{loess2d}, \texttt{MIC}, \texttt{TIC},
\texttt{spline}, and \texttt{stringy}. Combining with two jellyfish
hyper-parameters, a total of 42 cases is produced, comprising of 24
pipe-finding cases and 8 sine-wave finding cases. For each case, the JS
optimiser is run fifty times and the success rate is calculated as in
Section~\ref{sec-app-1}.

Smoothness and squintability are computed for each case, following the
procedures outlined in Section~\ref{sec-smoothness} and
Section~\ref{sec-squintability}. To calculate smoothness, three hundred
random bases are simulated. Index values and projection distance (to the
optimal basis) are calculated for each random basis before fitting them
into a Gaussian process model. For squintability, fifty random bases are
sampled and interpolated to the optimal basis with a step size of 0.005.
For these interpolated bases, index values and projection distances to
the optimal basis are calculated and averaged with a bin width of 0.005.
The squintability measure is then calculated from fitting a 4-parameter
scaled logistic function of index value against projection distance,
estimated by non-linear least squares.

Table~\ref{tbl-smoothness-squintability} presents the parameters
estimated from the Gaussian process (variance, range, smooth, and
nugget) and the scaled logistic function (\(\theta_1\) to \(\theta_4\))
for each case. The column ``smooth'' is used as the smoothness measure
and the column ``squint'' is calculated as {[}TODO: equation
reference{]} as squintability. The low squint value of MIC and TIC index
is due to its convex shape of the index value against the projection
distance, as shown in Figure~\ref{fig-idx-proj-dist}. Comparing to other
concave shapes, the maximum first derivative happens at a smaller
projection distance (\(\theta_2\)), requiring the optimiser to get
closer to the optimal to see a significant change in the index value,
hence a small squintability measure.

\caption{\label{fig-idx-proj-dist}Index values versus projection
distance for the 12 pipe/sine-wave finding problem, after the binning
procedure for calculating the squintability measure. The index values,
averaged at bin width of 0.005, are scaled from 0 to 1 for comparison
(holes, TIC, and stringy). The \texttt{MIC} and \texttt{TIC} index
curves are convex while others are concave. The stringy curve shows an
instantaneous jump to the optimum when aproaching the best basis.}

Table~\ref{tbl-mod-data} combines all the variables calculated above
(the JS optimiser success rate, jellyfish hyper-parameters, and
optimisation properties) into one table. Three preprocessing steps are
applied to the data: 1) scaling the jellyfish hyper-parameters by a
factor of 10 for interpretation, 2) creating a new binary variable
\texttt{long\_time} to indicate cases with an average run time over 30
seconds, and 3) encoding the success rate for the stringy case as 0,
since none of the 50 simulations identified the sine-wave shape. A
generalised linear model with a binomial family and a logit link
function is used to fit the data and Table~\ref{tbl-mod-output} presents
the model outputs. The model suggests that the success rate of the JS
optimiser is positively associated with the two jellyfish
hyper-parameters, as well as with smoothness and squintability. However,
being flagged with long runtime and an increase of data dimension will
reduce the success rate. The variable \texttt{squintability} and
\texttt{dimension} are significant, suggesting their importance relative
to jellyfish hyper-parameters in the optimisation success.

\begin{table}

\caption{\label{tbl-smoothness-squintability}Parameters estimated from
the Gaussian process (including variance, range, smoothness, and nugget)
and scaled logistic function (\(\theta_1\) to \(\theta_4\)) for the
pipe-finding and sine-wave finding problems. The column ``smooth'' and
``squint'' represent the smoothness and squintability measures.}

\centering\begingroup\fontsize{7}{9}\selectfont

\begin{tabular}{|>{}lll>{}r|rrr>{}r|rrr>{}r|>{}r|}
\toprule
  & shape & index & d & variance & range & smooth & nugget & $\theta_1$ & $\theta_2$ & $\theta_3$ & $\theta_4$ & squint\\
\midrule
1 & pipe & holes & 6 & 0.002 & 0.408 & 2.364 & 0.212 & 0.991 & 0.860 & 3.368 & 0.172 & 0.627\\
2 & pipe & holes & 8 & 0.000 & 0.259 & 2.373 & 0.613 & 1.011 & 0.869 & 3.264 & 0.085 & 0.696\\
3 & pipe & holes & 10 & 0.000 & 0.144 & 2.317 & 1.831 & 1.011 & 0.885 & 3.151 & 0.153 & 0.636\\
4 & pipe & holes & 12 & 0.000 & 0.254 & 2.173 & 0.879 & 0.999 & 0.878 & 3.345 & 0.196 & 0.621\\
5 & sine & MIC & 6 & 0.016 & 0.100 & 2.394 & 0.087 & 0.894 & 0.571 & 1.623 & -0.024 & 0.314\\
\addlinespace
6 & sine & MIC & 8 & 0.016 & 0.100 & 2.394 & 0.087 & 0.932 & 0.328 & 1.314 & -0.030 & 0.193\\
7 & sine & TIC & 6 & 0.124 & 0.104 & 2.471 & 0.086 & 0.951 & 0.536 & 1.719 & -0.025 & 0.330\\
8 & sine & TIC & 8 & 0.124 & 0.104 & 2.471 & 0.086 & 0.945 & 0.564 & 1.723 & -0.027 & 0.341\\
9 & sine & dcor2d & 6 & 0.034 & 0.167 & 2.663 & 0.114 & 0.954 & 1.039 & 2.742 & -0.019 & 0.737\\
10 & sine & loess2d & 6 & 0.083 & 0.307 & 2.194 & 0.292 & 1.016 & 1.039 & 2.648 & 0.080 & 0.689\\
\addlinespace
11 & sine & splines2d & 6 & 0.040 & 0.189 & 2.606 & 0.104 & 1.014 & 1.051 & 2.730 & -0.009 & 0.780\\
12 & sine & stringy & 6 & 0.000 & 1173.035 & 1.031 & 17608.047 & 1.040 & 0.011 & 254.733 & 0.170 & 0.648\\
\bottomrule
\end{tabular}
\endgroup{}

\end{table}%

@@

\toprule
Index & D & Smoothness & Squintability & Num. of jellyfish & Max. Num. of tries & Prop. of success & Time\\
\midrule
dcor2d\_2 & 6 & 2.663 & 0.737 & 20 & 50 & 0.44 & 35.320 secs\\
dcor2d\_2 & 6 & 2.663 & 0.737 & 20 & 100 & 0.60 & 70.347 secs\\
dcor2d\_2 & 6 & 2.663 & 0.737 & 50 & 50 & 0.90 & 87.158 secs\\
holes & 6 & 2.364 & 0.627 & 20 & 50 & 0.88 & 0.801 secs\\
holes & 6 & 2.364 & 0.627 & 20 & 100 & 0.96 & 1.456 secs\\
\addlinespace
holes & 6 & 2.364 & 0.627 & 50 & 50 & 1.00 & 1.477 secs\\
holes & 6 & 2.364 & 0.627 & 50 & 100 & 1.00 & 3.052 secs\\
\bottomrule

@@

\toprule
term & estimate & std.error & statistic & p.value\\
\midrule
Intercept & 2.685 & 15.464 & 0.174 & 0.862\\
Smoothness & 3.070 & 3.631 & 0.845 & 0.398\\
Squintability & -8.545 & 22.252 & -0.384 & 0.701\\
dimension (d) & -0.610 & 0.355 & -1.719 & 0.086\\
long time & -1.683 & 1.930 & -0.872 & 0.383\\
\addlinespace
number of jellyfish & 0.269 & 0.175 & 1.538 & 0.124\\
maximum number of tries & 0.080 & 0.196 & 0.410 & 0.682\\
\bottomrule

@@
